### 2024-10-11 teaching/math240f24/draft.tex
lecture oct 10

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}

@@

Next, suppose that the uniqueness
statement holds for $s$ (i.e., for any integer that decomposes
into a product of $s$ primes, the decomposition is unique) and we want to show that it holds
for $s+1$. We assume that
$$n = p_1p_2\cdots p_{s+1}$$
and
$$n = q_1q_2\cdots q_t,$$
where $p_1\le p_2\le \cdots \le p_{s+1}$ and $q_1\le q_2\le\cdots\le q_t$. We
have $p_{s+1}\divides n = q_1q_2\cdots q_t$, so by the preceding corollary, there
exists $1\le i\le t$ such that $p_{s+1}\divides q_i$, and since $q_i$ is prime, this means
that $p_{s+1} = q_i\le q_t$. Similarly, $q_t$ divides $n = q_1\ldots p_{s+1}$, so it
divides $p_j$ for some $1\le j\le s+1$. This means that $q_t = p_j \le p_{s+1}$. Hence
$q_t = p_{s+1}$. But
$$p_1p_2 \cdots p_s p_{s+1} = q_1q_2 \cdots q_t,$$
so we may divide out by $p_{s+1} = q_t$ on both sides to get
$${n\over p_{s+1}} = p_1p_2\cdots p_s = q_1 q_2 \cdots q_{t-1}.$$
By the induction hypothesis, the decomposition of $n/p_{s+1}$ is unique, so $t-1 = s$
and $p_i = q_i$ for all $1\le i\le t-1$. This proves that $s+1 = t$ and
$p_i = q_i$ for all $1\le i\le s+1$.\slug
\bye

We can use the Fundamental Theorem of Arithmetic to prove the following generalisation
of Theorem~{\thmsqrttwo}.

@@

\parenproclaim Theorem {\advthm} (Prime number theorem). For $x\in \RR$, let
$$\pi(x) = \bigl| \{ p \le x : p\ \hbox{prime}\} \bigr|.$$
Then $pi(x) \sim x/\ln x$, in the sense that
$$\lim_{x\to\infty} {\pi(x)\over x/\ln x} = 1.$$

@@

\advsect Modular Arithmetic

@@

\noindent\datestamp{14}{} {\it Proof.}\enspace
Suppose that $a\equiv_n b$, so that there exists $k\in \NN$ such that $a-b = kn$.
Then, let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
(So $r_1 = a\percent n$ and $r_2 = b\percent n$, and we have $0\le r_1,r_2 < n$.)
Write
$$kn = a - b = (q_1-q_2) n + (r_1 - r_2),$$
which we can rearrange to
$$r_1 - r_2 = kn - (q_1 - q_2)n = (k-q_1 + q_2)n.$$
This implies that $r_1 - r_2$ divides $n$, but
since $r_1, r_2\in [0,n)$, the quantity $r_1 - r_2$ is in the range $[-n-1, n)$. Hence
the only way it can divide $n$ is for $r_1 - r_2 = 0$. We conclude that $a\percent n = b\percent n$.

For the other direction, we once again
let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
Now the assumption is that $r_1 = a\percent n = b\percent n = r_2$. So
$$a - b = a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
This shows that $a\equiv b\pmod n$.\slug

This proposition is useful in practice. For example, suppose we want to know whether
$22\equiv 8\pmod 3$. We calculate $22 = 7\cdot 3 + 1$, so $22\percent 3 = 1$; meanwhile
$8 = 2\cdot 3 + 2$, so $8\percent 3 = 2$. By the proposition, this means that
$22\not\equiv 8\pmod 3$.

@@

For part (ii), suppose that $s,t\in \ZZ$ are such that $1 = sa+tn$. Then
taking this equation modulo $n$, we have
$$ 1 = sa+tn \equiv_n sa + t\cdot 0 \equiv_n sa.$$
Letting $b = s\percent n$, we have $1 = ba$, since $s$ and $b$ belong to the same
equivalence class modulo $n$, and $b$ is an integer in $[0,n-1]$ by the division algorithm.
This proves existence of such a $b$. To show uniqueness, suppose that $b$ and $c$ are both
integers in $[0,n-1]$ with $ab \equiv 1\pmod n$ and $ac\equiv 1\pmod n$. Then
$ab\equiv ac\pmod n$, and multiplying on both sides by $b$, we have $bab\equiv bac\pmod n$.
But $ba\equiv 1\pmod n$, so cancelling this out in the equation, we have $b\equiv c\pmod n$.
In other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
and the only divisor of $n$ in this range is $0$. Hence $b-c = 0$ and we conclude that $b=c$.\slug

@@

\noindent\datestamp{14}{} \hskip\parindent
We now investigate what happens in the particular case where we the modulus $n$ is prime.
In this situation we have the following proposition.

@@

\proclaim Lemma \advthm. For all prime numbers $p$, the integer $(p-1)!$ is congruent to $-1$ modulo $p$.

@@

@@

\bye

>>>>

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}

@@

Next, suppose that the uniqueness
statement holds for $s$ (i.e., for any integer that decomposes
into a product of $s$ primes, the decomposition is unique) and we want to show that it holds
for $s+1$. We assume that
$$n = p_1p_2\cdots p_{s+1}$$
and
$$n = q_1q_2\cdots q_t,$$
where $p_1\le p_2\le \cdots \le p_{s+1}$ and $q_1\le q_2\le\cdots\le q_t$. We
have $p_{s+1}\divides n = q_1q_2\cdots q_t$, so by the preceding corollary, there
exists $1\le i\le t$ such that $p_{s+1}\divides q_i$, and since $q_i$ is prime, this means
that $p_{s+1} = q_i\le q_t$. Similarly, $q_t$ divides $n = q_1\ldots p_{s+1}$, so it
divides $p_j$ for some $1\le j\le s+1$. This means that $q_t = p_j \le p_{s+1}$. Hence
$q_t = p_{s+1}$. But
$$p_1p_2 \cdots p_s p_{s+1} = q_1q_2 \cdots q_t,$$
so we may divide out by $p_{s+1} = q_t$ on both sides to get
$${n\over p_{s+1}} = p_1p_2\cdots p_s = q_1 q_2 \cdots q_{t-1}.$$
By the induction hypothesis, the decomposition of $n/p_{s+1}$ is unique, so $t-1 = s$
and $p_i = q_i$ for all $1\le i\le t-1$. This proves that $s+1 = t$ and
$p_i = q_i$ for all $1\le i\le s+1$.\slug

\noindent\datestamp{10}{x} \hskip\parindent
We can use the Fundamental Theorem of Arithmetic to prove the following generalisation
of Theorem~{\thmsqrttwo}.

@@

\parenproclaim Theorem {\advthm} (Prime number theorem). For $x\in \RR$, let
$$\pi(x) = \bigl| \{ p \le x : p\ \hbox{prime}\} \bigr|.$$
Then $\pi(x) \sim x/\ln x$, in the sense that
$$\lim_{x\to\infty} {\pi(x)\over x/\ln x} = 1.$$

@@

\advsect Modular arithmetic

@@

\proof
Suppose that $a\equiv_n b$, so that there exists $k\in \NN$ such that $a-b = kn$.
Then, let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
(So $r_1 = a\percent n$ and $r_2 = b\percent n$, and we have $0\le r_1,r_2 < n$.)
Write
$$kn = a - b = (q_1-q_2) n + (r_1 - r_2),$$
which we can rearrange to
$$r_1 - r_2 = kn - (q_1 - q_2)n = (k-q_1 + q_2)n.$$
This implies that $r_1 - r_2$ divides $n$, but
since $r_1, r_2\in [0,n)$, the quantity $r_1 - r_2$ is in the range $[-n-1, n)$. Hence
the only way it can divide $n$ is for $r_1 - r_2 = 0$. We conclude that $a\percent n = b\percent n$.

For the other direction, we once again
let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
Now the assumption is that $r_1 = a\percent n = b\percent n = r_2$. So
$$a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
This shows that $a\equiv b\pmod n$.\slug
\bye

\noindent\datestamp{14}{} \hskip\parindent
This proposition is useful in practice. For example, suppose we want to know whether
$22\equiv 8\pmod 3$. We calculate $22 = 7\cdot 3 + 1$, so $22\percent 3 = 1$; meanwhile
$8 = 2\cdot 3 + 2$, so $8\percent 3 = 2$. By the proposition, this means that
$22\not\equiv 8\pmod 3$.

@@

For part (ii), suppose that $s,t\in \ZZ$ are such that $1 = sa+tn$. Then
taking this equation modulo $n$, we have
$$ 1 = sa+tn \equiv_n sa + 0t \equiv_n sa.$$
Letting $b = s\percent n$, we have $1 = ba$, since $s$ and $b$ belong to the same
equivalence class modulo $n$, and $b$ is an integer in $[0,n-1]$ by the division algorithm.
This proves existence of such a $b$. To show uniqueness, suppose that $b$ and $c$ are both
integers in $[0,n-1]$ with $ab \equiv 1\pmod n$ and $ac\equiv 1\pmod n$. Then
$ab\equiv ac\pmod n$, and multiplying on both sides by $b$, we have $bab\equiv bac\pmod n$.
But $ba\equiv 1\pmod n$, so cancelling this out in the equation, we have $b\equiv c\pmod n$.
In other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
and the only divisor of $n$ in this range is $0$. Hence $b-c = 0$ and we conclude that $b=c$.\slug

@@

\noindent\datestamp{15}{} \hskip\parindent
We now investigate what happens in the particular case where we the modulus $n$ is prime.
In this situation we have the following proposition.

@@

\edef\lemhalfwilson{\the\thmcount}
\proclaim Lemma \advthm. For all prime numbers $p$, the integer $(p-1)!$ is congruent to $-1$ modulo $p$.

@@

As a matter of interest, the intermediary lemma we proved is one direction of Wilson's theorem, proved
by John Wilson in 1770.

\parenproclaim Theorem {\advthm} (Wilson's theorem).
For all integers $n\ge 2$, the congruence
$$(n-1)!\equiv -1\pmod n$$
holds if and only if $n$ is prime.

\proof We already proved the ``if'' direction earlier, as Lemma~{\lemhalfwilson}.
We leave the ``only if'' direction of the proof as an exercise for the reader.\slug

@@

\advsect Applications of number theory

In this section we present a potpourri of interesting ways number theory is applied to make your life better.

\medskip\boldlabel {\ninepoint\bf ISBN} book identifiers. Every published book has an
{\mc ISBN} code that serves as its
unique identifier. This code is $10$ digits long for books published before 2007, but three new digits
have been added for books published after 2007. In this section we'll deal with the simpler case of $10$ digits.

What we want to do is to bake some redundancy into the codes, so that if a single digit is typed wrong,
a computer will be able to tell the user that the code is invalid. This allows the user to correct their
mistake. The way we do this, in a ten-digit code $d_{10}d_9d_8\cdots d_2d_1$,
is to have only the first nine digits encode the book. The last digit is a {\it check digit},
chosen so that
\edef\eqvalidisbn{\the\eqcount}
$$\sum_{i=1}^{10} i d_i \equiv 0 \pmod{11}.\adveq$$
In other words,
$$d_1\equiv -\sum_{i=2}^{10} id_i \pmod{11}.$$
If $d_1$ needs to equal $10$, we use the symbol `X'.

How does this solve our problem? Well, if a hapless person takes a valid {\mc ISBN} code
and mangles it by either getting one digit wrong, or by swapping two adjacent digits,
the result is an invalid {\mc ISBN} code, and the computer will be able to flag it.
We formalise this statement as the following theorem.

\proclaim Theorem \advthm. Let $d_{10} d_9 d_8\cdots d_2 d_1$ be a valid {\mc ISBN}
code; that is, a code that satisfies~\refeq{\eqvalidisbn}. Then any code obtained by either
\medskip
\item{i)} changing exactly one digit $d_i$, for some $1\le i\le 10$, or
\smallskip
\item{ii)} swapping {\it distinct} adjacent digits $d_i$ and $d_{i-1}$, for some $2\le i\le 10$,
\medskip
is not a valid {\mc ISBN} codes (it does not satisfy~\refeq{\eqvalidisbn}).

\proof First we prove (i). Suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
equal to $d_{10} d_9 d_8\cdots d_2 d_1$ except at one place $1\le j\le 10$. So $c_j\ne d_j$,
but for all $1\le i\le 10$ with $i\ne j$, we have $c_i = d_i$. Then
$$\eqalign{
\sum_{i=1}^{10} i c_i &\equiv_11 \sum_{i=1}^{10} i c_i - 0 \cr
&\equiv_11 \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
&= \sum_{i=1}^{10} i(c_i - d_i) \cr
&= j(c_j-d_j). \cr
}$$
But $j\in \{1,\ldots, 10\}$ and $(c_j-d_j) \in [-10, 10]$. So neither $j$ nor $c_j-d_j$
are congruent to $0$ modulo $11$. Since $11$ is prime, their product cannot be congruent to $0$
modulo $11$. We conclude that $\sum_{i=1}^{10} i c_i\not\equiv 0\pmod 11$; i.e., this is
not a valid {\mc ISBN} code.

To prove part (ii), suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
equal to $d_{10} d_9 d_8\cdots d_2 d_1$, except that for some $2\le j\le 10$, we
have $c_j = d_{j-1}$ and $c_{j-1}=d_j$ (the adjacent digits are swapped). Furthermore,
assume that $d_j\ne d_{j-1}$, since swapping adjacent digits that are the same doesn't
do anything to the code. Then
$$\eqalign{
\sum_{i=1}^{10} i c_i &\equiv_11 \sum_{i=1}^{10} i c_i - 0 \cr
&\equiv_11 \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
&= \sum_{i=1}^{10} i(c_i - d_i) \cr
&= (j-1)c_{j-1} + jc_j - (j-1)d_{j-1} - jd_j \cr
&= (j-1)d_j + jd_{j-1} - (j-1)d_{j-1} - jd_j \cr
&= d_{j-1} - d_j.\cr
}$$
But since $d_{j-1}$ and $d_j$ are both in the range $\{0,\ldots, 10\}$, their difference
is in the range $[-10,10]$, and we assumed that $d_{j-1}\ne d_j$, so this difference is not zero.
We conclude that $d_{j-1}-d_j\not\equiv 0\pmod 11$, so the code is not valid.\slug

Something similar is done with credit card numbers, though we don't take digits modulo $11$
(or else we'd have to use the digit 'X', which is cumbersome).
Instead, a slightly different method called Luhn's algorithm is used. The calculation
used to find the check digit is a bit different (let's not get into it),
but the result is that we can still detect
single-digit errors, and we can detect all swapping errors except for the transposition
$09\leftrightarrow 90$.

\medskip\boldlabel Divisibility tests.
In grade school you might have learned the following divisiblity rule:
{\sl To tell if a number is divisible by $3$, you sum up all of its digits. If this is a
multiple of $3$, then the original number was a multiple of $3$, and if not, then the original
number wasn't.} The same holds with $3$ replaced by $9$.

Using the number theory we've learned, we are now able to prove this fact.

\proclaim Proposition \advthm. Suppose that $n\in \NN$ is written in base-$10$ digits as
$$n = d_k d_{k-1} d_{k-2}\cdots d_2 d_1 d_0,$$
for some integer $k\ge 0$ and $d_i\in \{0,\ldots,9\}$ for $0\le i\le k$.
Then $n$ is divisible by $3$ if and only if
$$\sum_{i=0}^k d_i$$
is also divisible by $3$.

\proof Since $10\equiv 1\pmod 3$, any power of $10$ is also congruent to $1$ modulo $3$. So we have
$$n = \sum_{i=0}^k d_i 10^i \equiv_3 \sum_{i=0}^k d_i,$$
and we see that $n$ is divisible by $3$ if and only if the sum on the right-hand side is.\slug

This works with $3$ replaced by $9$ since we also have $10\equiv 1\pmod 9$.
